SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-12-07 design_arch.tex
Updating Pval Image, Incorporating suggestions in Design

\pg{the text below would be easier to follow if you used the figure
  more. make reference to the figure at the points that you are
  talking about}

\noindent \textbf{Inferencing}\\
%ETG:is this like an explanation of how prefetching happens? 
%EB: I see this as a short context of the datapath leading to the perc filter.
The base prefetcher is triggered on every demand access to the L2
Cache.  At this point, it has the opportunity to trigger a prefetch.
If it does so, it will also need to decide how many cache blocks to
prefetch.  These blocks can be either placed in the L2 or L3 cache
according to the confidence of the prefetching mechanism.  Once the
base prefetcher is triggered, the suggested prefetch candidates are
fed to the perceptron filter to determine the usefulness of these
prefetches. The filter ultimately decides whether to issue the
prefetch suggestions of the base prefetcher. To make the decision,
each feature corresponding to a suggested prefetch is used to index a
table and all the corresponding weights are summed. The sum denotes
the confidence value for the suggested prefetch, and is thresholded
against two different magnitudes: $\tau_{hi}$ and $\tau_{lo}$.

@@

\noindent \textbf{Recording}\\
As shown in step 2 of Figure~\ref{fig:PPF_Datapath} the prefetches
that make it through the inference stage are recorded in the
``Prefetch Table''. The prefetch table is a 1,024-entry, direct mapped
structure that contains all metadata required to re-index the
perceptron entries for training.

@@

\noindent \textbf{Feedback and Data Retrieval}\\
When there is an eviction or a demand access to the L2, training for
both the base prefetcher and our filter mechanism is triggered. The
address of the cache block that triggered training is used to index
both the Prefetch and Reject tables. 10 bits of the address are used
to index tables, and another 6 bits to perform tag matching. Once the
base prefetcher has suggested a prefetch request, the corresponding
features are used to index the tables of weights.

@@

\noindent \textbf{Training}\newline The address from the demand
request triggering the training is looked up in both tables. If the
address is in the prefetch table and marked as valid, this hints the
previous prediction was correct and this is a useful prefetch. We
compute the sum of the corresponding weights. If the sum falls below a
specific threshold, training occurs and the corresponding weights are
adjusted accordingly. These thresholds are introduced in order to
avoid over-training, helping the filter adapt quickly to changes in
memory behavior. These thresholds are referred to as $\theta_p$ and
$\theta_n$, respectively for the positive and negative values of
training saturation.

@@

Parallel to accessing the prefetch table, on a demand access, the
reject table is accessed. Before the demand access triggers the next
set of prefetches, the reject table is checked for a valid entry. A
hit means that the corresponding cache block was initially suggested
by the base prefetcher, but wrongly rejected by the perceptron
filter. The perceptron filter learns from this and makes use of the
corresponding features associated to the original prefetch request,
which are stored in the reject table, to index the weights tables and
adjust the weights accordingly.

\pg{what prior work? this reference is too oblique to be useful,
  either be specific or drop this paragraph} Prior work in this domain
lacks the knowledge of which cache blocks should have been prefetched,
but were not. The implementation of the reject table, allows us to
have this information, which can be use to further optimize our
prefetching mechanism.

@@

\subsection{Optimizing PPF for a Given Prefetcher}
\label{Arch-Generalizing}
The above discussion of PPF shows that it is highly modular and can be
adapted to be used over any base prefetcher for increased prefetch
accuracy.  Only two hooks are required between PPF and the baseline
prefetcher.  The first is to make sure that all the prefetch
candidates of the prefetcher pass through the perceptron filter and if
qualified, the metadata for perceptron indexing be stored. The second
is needed when the feedback of a prior prefetch is available in form
of a subsequent demand hit or cache eviction. In that case, the stored
metadata needs to be retrieved to update the state of the perceptrons.
\pg{this isn't quite right, I thought we were taking some metadata
  from the prefetcher itself like confidence.  You should explicitly
  mention that prefetcher metadata can be used, update the figure to
  show that metadata passing from SPP to the prefetcher and point to
  the figure to show where that connection is}

@@

\noindent \textbf{Make Base Prefetcher More Aggressive:} Tune down any
internal thresholds or throttling mechanisms to increase its
aggressiveness.

@@

>>>>

%%\pg{the text below would be easier to follow if you used the figure
%%  more. make reference to the figure at the points that you are
%%  talking about}

\noindent \textbf{Inferencing}\\
%ETG:is this like an explanation of how prefetching happens? 
%EB: I see this as a short context of the datapath leading to the perc filter.
The base prefetcher is triggered on every demand access to the L2
Cache, as seen in Figure~\ref{fig:PPF_Hierarchy}.  At this point, 
it has the opportunity to trigger a prefetch.
If it does so, it will also need to decide how many cache blocks to
prefetch.  These blocks can be either placed in the L2 or L3 cache
according to the confidence of the prefetching mechanism.  Once the
base prefetcher is triggered, the suggested prefetch candidates are
fed to the perceptron filter to determine the usefulness of these
prefetches. The filter ultimately decides whether to issue the
prefetch suggestions of the base prefetcher. As shown in step 1
of Figure~\ref{fig:PPF_Datapath}(a), to make the decision,
each feature corresponding to a suggested prefetch is used to index a
table and all the corresponding weights are summed. The sum denotes
the confidence value for the suggested prefetch, and is thresholded
against two different values: $\tau_{hi}$ and $\tau_{lo}$.

@@

\noindent \textbf{Recording}\\
As shown in step 2 of Figure~\ref{fig:PPF_Datapath}(a) the prefetches
that make it through the inference stage are recorded in the
``Prefetch Table''. The prefetch table is a 1,024-entry, direct mapped
structure that contains all metadata required to re-index the
perceptron entries for training.

@@

\noindent \textbf{Feedback and Data Retrieval}\\
As depicted in step 3, when there is an eviction or a demand access 
to the L2, training for
both the base prefetcher and our filter mechanism is triggered. The
address of the cache block that triggered training is used to index
both the Prefetch and Reject tables. 10 bits of the address are used
to index tables, and another 6 bits to perform tag matching. Once the
base prefetcher has suggested a prefetch request, the corresponding
features are used to index the tables of weights.

@@

\noindent \textbf{Training}\newline As can be seen in step 4 of, 
the address from the demand
request triggering the training is looked up in both tables. If the
address is in the prefetch table and marked as valid, this hints the
previous prediction was correct and this is a useful prefetch. We
compute the sum of the corresponding weights. If the sum falls below a
specific threshold, training occurs and the corresponding weights are
adjusted accordingly. These thresholds are introduced in order to
avoid over-training, helping the filter adapt quickly to changes in
memory behavior. These thresholds are referred to as $\theta_p$ and
$\theta_n$, respectively for the positive and negative values of
training saturation.

@@

Parallel to accessing the prefetch table, on a demand access, the
reject table is accessed. Before the demand access triggers the next
set of prefetches, the reject table is checked for a valid entry. A
hit means that the corresponding cache block was initially suggested
by the base prefetcher, but wrongly rejected by the perceptron
filter. The perceptron filter learns from this and makes use of the
corresponding features associated to the original prefetch request,
which are stored in the reject table, to index the weights tables and
adjust the weights accordingly.
The implementation of the reject table, allows us to capture the 
information of prefetches that were rejected, and that can be used 
to further optimize our prefetching mechanism.
%
%\pg{what prior work? this reference is too oblique to be useful,
%  either be specific or drop this paragraph} 
% Prior work in this domain
% lacks the knowledge of which cache blocks should have been prefetched,
% but were not.

@@

\subsection{Optimizing PPF for a Given Prefetcher}
\label{Arch-Generalizing}
The above discussion of PPF shows that it is highly modular and can be
adapted to be used over any base prefetcher for increased prefetch
accuracy. As a first step, all the prefetch
candidates of the prefetcher have to pass through the perceptron filter. 
If qualified, the metadata for perceptron indexing has to be stored. 
Next, when the feedback of a prior prefetch is available in form
of a subsequent demand hit or cache eviction, the stored
metadata needs to be retrieved to update the state of the perceptrons.
% \pg{this isn't quite right, I thought we were taking some metadata
%   from the prefetcher itself like confidence.  You should explicitly
%   mention that prefetcher metadata can be used, update the figure to
%   show that metadata passing from SPP to the prefetcher and point to
%   the figure to show where that connection is}

@@

\noindent \textbf{Making Base Prefetcher More Aggressive:} By Tuning 
down any internal thresholds or throttling mechanisms to increase its
aggressiveness.

@@

\noindent \textbf{Using Metadata from the Prefetcher:} Some of the 
internal counters specific to the underlying prefetcher can be suitable candidates
for the perceptron features. To make sure that the perceptron layer sees that, 
the relevant meta-data must be exported from the prefetcher to PPF. This way,
PPF can be optimized to work tightly-knit with the base prefetcher.

### 2018-12-07 design_impl.tex
Updating Pval Image, Incorporating suggestions in Design

\noindent \textbf{Original Thresholds Discarded:} In PPF, the
perceptron sum is used to decide whether to prefetch or not, and the
fill-level in case of prefetch.  Thus, the confidence thresholds --
T\textsubscript{f} and T\textsubscript{p} are no longer needed to
throttle the prefetcher.  \pg{Note, however, as we discuss later, we
  use these confidence thresholds a feature in PPF so the mechanism is
  retained. -- right?}
% djimenez: keeping comments on every line here so it doesn't screw up my
%"roll-your-own" itemized list. don't put in any newlines.
%
%
% EB: Can this point be removed?
% djimenez: I think this would be necessary to reproduce our results so leave it in. 
% EB: It's ultimately a check to avoid SPP's internal segfault. Won't have a real impact.
%
%\item \textbf{Looking at L2 MSHR Queue while prefetching}\newline The
%  original SPP does not consider the slots available in the L2 MSHR
%  queue.  It has the assumption that at no time can the number of
%  suggested prefetches exceed the capacity of the queue.  While it was
%  not described in the paper, the internal confidence mechanism makes
%  sure that the above assumption was maintained.  In PPF, since there
%  cannot be an assumption on the degree of aggressiveness of the
%  underlying prefetcher, we explicitly check that at no point should
%  the number of suggested prefetches exceed the L2 MSHR queue. \pg{is
%    this a big deal?  Seems like a detail}
%

\pg{since you actually moved this table into PPF, most this discussion
  should actually go in the PPF arch section, I think its generic to
  all PPFs likely anyways.  Instead here I'd talk about how you
  removed the prefetch filter that was part of SPP since its redundant
  with the function of PPF}
\noindent \textbf{Enhanced Prefetch Table:} The prefetch table
introduced in the original SPP keeps track of up to 1024 L2 cache
prefetch suggestions. For PPF, the prefetch table is modified to store
the features used for perceptron indexing.  This metadata is needed to
retrieve the values of the features to train the perceptron at a later
stage in time. The prefetch table mentioned above keeps track of the
prefetches that actually happened in the L2 Cache.  We introduce the
reject table to keep a track of the prefetches that were suggested by
the base SPP engine but were rejected by PPF. In case it is detected
that a future demand fetch could have used this rejected prefetch, the
perceptron weights are updated to reflect that misprediction.

@@

\noindent \textbf{Base Address}: The base address of the demand access
that triggers the prefetch corresponds to a stream of future accesses
that SPP and the PPF have seen before. Therefore, PPF will correlate
the past behavior of this address to the confidence of a
prefetch. \pg{what do you mean by ``base address'' the full physical
  address?  Be specific.  Which bits and why? You should probably also
  talk about what you do when the number of bits is unwieldy.}

>>>>

\noindent \textbf{Exporting Features from SPP:} PPF uses the meta-data 
specific to SPP, to build some of the perceptron features. These include
the lookahead depth, signature and the confidence counter. These features 
were made visible to PPF\\
%
\noindent \textbf{Original Thresholds Discarded:} In PPF, the
perceptron sum is used to decide whether to prefetch or not, and the
fill-level in case of prefetch.  Thus, the confidence thresholds used 
by SPP -- T\textsubscript{f} and T\textsubscript{p} are no longer needed to
throttle the prefetcher and can be discarded.
%
%\pg{Note, however, as we discuss later, we
%use these confidence thresholds a feature in PPF so the mechanism is
%retained. -- right?}
% djimenez: keeping comments on every line here so it doesn't screw up my
%"roll-your-own" itemized list. don't put in any newlines.
%
%
% EB: Can this point be removed?
% djimenez: I think this would be necessary to reproduce our results so leave it in. 
% EB: It's ultimately a check to avoid SPP's internal segfault. Won't have a real impact.
%
%\item \textbf{Looking at L2 MSHR Queue while prefetching}\newline The
%  original SPP does not consider the slots available in the L2 MSHR
%  queue.  It has the assumption that at no time can the number of
%  suggested prefetches exceed the capacity of the queue.  While it was
%  not described in the paper, the internal confidence mechanism makes
%  sure that the above assumption was maintained.  In PPF, since there
%  cannot be an assumption on the degree of aggressiveness of the
%  underlying prefetcher, we explicitly check that at no point should
%  the number of suggested prefetches exceed the L2 MSHR queue. \pg{is
%    this a big deal?  Seems like a detail}
%

%% \pg{since you actually moved this table into PPF, most this discussion
%%   should actually go in the PPF arch section, I think its generic to
%%   all PPFs likely anyways.  Instead here I'd talk about how you
%%   removed the prefetch filter that was part of SPP since its redundant
%%   with the function of PPF}
%% \noindent \textbf{Enhanced Prefetch Table:} The prefetch table
%% introduced in the original SPP keeps track of up to 1024 L2 cache
%% prefetch suggestions. For PPF, the prefetch table is modified to store
%% the features used for perceptron indexing.  This metadata is needed to
%% retrieve the values of the features to train the perceptron at a later
%% stage in time. The prefetch table mentioned above keeps track of the
%% prefetches that actually happened in the L2 Cache.  We introduce the
%% reject table to keep a track of the prefetches that were suggested by
%% the base SPP engine but were rejected by PPF. In case it is detected
%% that a future demand fetch could have used this rejected prefetch, the
%% perceptron weights are updated to reflect that misprediction.

@@

\noindent \textbf{Physical Address}: Here we use the lower bits 
of the physical address of 
the demand access that triggers the prefetch. This address corresponds 
to a  stream of accesses that SPP and the PPF have seen before. 
Therefore, PPF will correlate the past behavior of this address 
to the prefetch outcome. 
%
%\pg{what do you mean by ``base address'' the full physical
%  address?  Be specific.  Which bits and why? You should probably also
%  talk about what you do when the number of bits is unwieldy.}

### 2018-12-07 method.tex
fix link
Updating Pval Image, Incorporating suggestions in Design

%This is the first time that SPEC CPU 2017 benchmark suite~\cite{SPEC2017}
%has been used to characterize and measure the prefetch performance.
%We use the SPEC CPU 2017 benchmark suite~\cite{SPEC2017}. 
We use all the 20 workloads available in the SPEC CPU 2017 suite~\cite{SPEC 2017}. Using the SimPoint~\cite{SimPoint}
methodology, we identified 95 different program segments of 1 Billion
instructions each.\\
%
\noindent \textit{Single-core performance:} For single-core simulations, we use 200
million instructions to warm-up the microarchitectural structures and the next
one billion instructions to do detailed simulations and collect run-time
statistics. We report the IPC speedup over the baseline of no prefetching.
The final numbers reported are the geometric mean of the weighted mean speedup 
achieved per application using the SimPoint methodology.\\
%
\noindent \textit{Multi-core performance:} For multi-application workloads, we generate
100 random mixes and another 100 mixes from the memory intensive subset of
SPEC CPU 2017. For 4-core workload, 200 Million instructions are used for warm-up
and additional 1 Billion instruction simulated for collecting statistics.
Each CPU keeps executing its workload till the last CPU completes one billion
instructions after warm-up. For collecting IPC and other data, only the first
billion instructions are considered as the region of interest.

@@

\subsection{Developing Features for PPF}
\label{Method-Features}
This section describes the intuition and analysis that went behind developing
the features. As noted earlier, we developed a set of nine features that allow
the perceptron layer to correlate prefetching decision with the program
context. To study the correlation across each feature, we study statistically
the perceptron weights and try to interpret their distribution. \\ \\
%
\textbf{Global Pearson's Correlation}\newline 
For this experiment, we examine
the perceptron weights at the end of all trace execution by which time the
weights have settled to steady values. The weights obtained from running all
the SPEC CPU 2017 traces are concatenated. The feature with bulk of the
perceptron weights concentrated around 0 or small magnitude numbers show a
weak correlation with the prefetching outcome. On the other hand, features
with most of the weights saturated around highest value (+15) show a high
positive correlation and the features with weights close to the lowest value
(-16) show a strong negative correlation.

@@

Figure~\ref{Fig:Pvals} shows all the features which are finally used, 
arranged in the increasing order of their Pearson's factor. 
As can be seen 5 out of the 9 features provide a moderate to high 
correlation, with the magnitude of P-value > 0.6. The single most 
important feature, \textit{Confidence XOR Page address} helps provide
a correlation to prefetch outcome with a factor of 0.90. \\ \\
%
\textbf{Per Trace Correlation} \newline Another important way to look
at the perceptron features is to see how much their contribution
varies across the traces. Here we give special attention to features
with low P-values in. Figure~\ref{Fig:Pvals_Trace} shows the
variation of P-values for three features : \textit{PC XOR Delta},
\textit{Signature XOR Delta} and 
\textit{PC XOR Depth}; across all the SPEC CPU 2017 traces. 
For simplicity, the traces are arranged in an
increasing order of contribution made by the feature. 
It can be seen that even features with a low
overall correlation provide useful correlation (magnitude > 0.5) for
a significant number of traces. \\ \\
%
\textbf{Trimming Features Using Cross-correlation} \newline Beside
providing interesting insights into prefetching behavior, P-value can
also be used for feature selection and prefetcher tuning.

@@

We initially came up a set of 23 features. By studying
cross correlation of each of these features against others in a 23$\times$23
matrix, we identified pairs of features with correlation factor > 0.9 in
magnitude and eliminated redundant features, reducing the feature count to 9.
Thus, in the final implementation of PPF, no two features have a high
correlation between them. This way we can be sure that each feature makes a
contribution that cannot be captured using other features.

>>>>

%This is the first time that SPEC CPU 2017 benchmark suite~\cite{SPEC2017}
%has been used to characterize and measure the prefetch performance.
%We use the SPEC CPU 2017 benchmark suite~\cite{SPEC2017}. 
We use all the 20 workloads available in the SPEC CPU 2017 suite~\cite{SPEC2017}. Using the SimPoint~\cite{SimPoint}
methodology, we identified 95 different program segments of 1 Billion
instructions each.\\
%
\noindent \textit{Single-core performance:} For single-core simulations, we use 200
million instructions to warm-up the microarchitectural structures and the next
one billion instructions to do detailed simulations and collect run-time
statistics. We report the IPC speedup over the baseline of no prefetching.
The final numbers reported are the geometric mean of the weighted mean speedup 
achieved per application using the SimPoint methodology.\\
%
\noindent \textit{Multi-core performance:} For multi-application workloads, we generate
100 random mixes and another 100 mixes from the memory intensive subset of
SPEC CPU 2017. For 4-core workload, 200 Million instructions are used for warm-up
and additional 1 Billion instruction simulated for collecting statistics.
Each CPU keeps executing its workload till the last CPU completes one billion
instructions after warm-up. For collecting IPC and other data, only the first
billion instructions are considered as the region of interest.

@@

\subsection{Developing Features for PPF}
\label{Method-Features}
This section describes the intuition and analysis that went behind developing
the perceptron features. As noted earlier, we developed a set of 
nine features that allows the perceptron layer to correlate prefetching decision 
with the program behaviour. To study the correlation across each feature, 
we statistically examine the perceptron weights and try to 
interpret their distribution. \\ \\
%
\textbf{Global Pearson's Correlation}\newline 
Here we examine
the perceptron weights at the end of all trace execution by which time the
weights have settled to steady values. The weights obtained from running all
the SPEC CPU 2017 traces are concatenated. The feature with bulk of the
perceptron weights concentrated around 0 or small magnitude numbers show a
weak correlation with the prefetching outcome. On the other hand, features
with most of the weights saturated around highest value (+15) show a high
positive correlation and the features with weights close to the lowest value
(-16) show a strong negative correlation.

@@

Figure~\ref{Fig:Pvals} shows all the features which are finally used, 
arranged in the increasing order of their Pearson's factor. 
As can be seen 5 out of the 9 features provide a moderate to high 
correlation, with the magnitude of P-value > 0.6. The single most 
important feature, \textit{Confidence XOR Page address} helps provide
a correlation to prefetch outcome with a factor of 0.90. \\ \\
%
\textbf{Per Trace Correlation} \newline Another important way to look
at the perceptron features is to see how much their contribution
varies across the traces. Here we give special attention to features
with low P-values in. Figure~\ref{Fig:Pvals_Trace} shows the
variation of P-values for three features : \textit{PC XOR Delta},
\textit{Signature XOR Delta} and 
\textit{PC XOR Depth}; across all the SPEC CPU 2017 traces. 
For simplicity, the traces are arranged in an
increasing order of contribution made by the feature. 
It can be seen that even features with a low
overall correlation provide useful correlation (magnitude > 0.5) for
a significant number of traces. This study motivated us to choose \textit{PC XOR DELTA}
over \textit{Last Signature} as it provided useful correlation in at least some of the
traces.\\ \\
%
\textbf{Trimming Features Using Cross-correlation} \newline Beside
providing interesting insights into prefetching behavior, P-value can
also be used for feature selection and prefetcher tuning.

@@

We initially came up a set of 23 features. By studying
cross correlation of each of these features against others in a 23$\times$23
matrix, we identified pairs of features with correlation factor > 0.9 in
magnitude and eliminated redundant features, using guidance from Global and per-trace
Pearson's factor of those features. By doing this, we reduced the feature count to 9.
Thus, in the final implementation of PPF, no two features have a high
correlation between them. This way we can be sure that each feature makes a
contribution that cannot be captured using other features.
